---
title: "2480 Final Project"
author: "Yi Yang"
date: "2024-04-11"
output:
  pdf_document: default
  html_document: default
---
Package Upload
```{r, message=FALSE, warning=FALSE}
library(readr)
library(haven)
library(psych)
library(tidyverse)
library(labelled)
library(table1)
library(dplyr)
library(haven)
library(tidyverse)
library(ggplot2)
library(lme4)
library(broom)
library(naniar)
library(sjPlot)
library(labelled)
library(performance)
library(knitr)
library(kableExtra)
library(lmerTest)
library(pander)
library(performance)
library(corrplot)
```
### Upload Data
```{r}
data <- read_dta("finalproj_2023.dta")
head(data)
```
### Data Cleaning
```{r}
data <- data %>%
 mutate(PID = (ER30001 * 1000) + ER30002) %>%
 relocate(PID) #putting at beginning of dataset
obs <- dim(data)[1]
obs

sum(duplicated(data$PID))


```

```{r}
data$PID <- as.integer(data$PID)
data$Anxiety1<- data$TA050933
data$Anxiety2<- data$TA070914
data$Anxiety3<- data$TA090978
data$Anxiety4<- data$TA111120
data$Anxiety5<- data$TA131212
data$Smoking <- data$TA050762
data$Race <- data$TA050884
data$age1 <- data$ER33804
data$age2 <- data$ER33904
data$age3 <- data$ER34004
data$age4 <- data$ER34104
data$age5 <- data$ER34204


sample_dat <- data %>%
  select(Anxiety1,Anxiety2,Anxiety3,Anxiety4,Anxiety5,
         Smoking, Race, age1,age2,age3,age4,age5,PID) %>%
  dplyr::mutate(Race = case_when(
    Race == 1 ~ "White",
    Race == 2 ~ "Black",
    Race == 3 ~ "Other",
    Race == 4 ~ "Other",
    Race == 5 ~ "Other", 
    Race == 7 | Race == 8 | Race == 9 ~ NA_character_
  )) 
table(data$Race, useNA = "always")
sample_dat_1 <- sample_dat %>% filter(!is.na(Anxiety1))
dim(sample_dat_1)
sample_dat_2 <- sample_dat_1 %>% filter(!is.na(Anxiety2))
sample_dat_3 <- sample_dat_2 %>% filter(!is.na(Anxiety3))
sample_dat_4 <- sample_dat_3 %>% filter(!is.na(Anxiety4))
sample_dat_5 <- sample_dat_4 %>% filter(!is.na(Anxiety5))
sample_dat_6 <- sample_dat_5 %>% filter(!is.na(Smoking))
dim(sample_dat_6)


table1(~.|Anxiety1 , data = sample_dat_6)
```


### 3.a Descriptive Statistics of the data
```{r}
head(sample_dat_6)
dim(sample_dat_6)

des<-sample_dat_6 %>% describe()
# Descriptive statistics of the data
des

# Calculate the correlation coefficient matrix for Anxiety
cor_anx<-cor(sample_dat_6[c("Anxiety1","Anxiety2","Anxiety3","Anxiety4", "Anxiety5")], 
             use = "pairwise.complete.obs" )
# correlation coefficient matrix for Anxiety
cor_anx
```

### 3.b i Describe the growth in your outcome
```{r, warning=FALSE}
obs <- dim(sample_dat_6)[1] # data size
set.seed(0)
sample_data <- sample_dat_6[sample(obs, size = 25),] # Sampling of 25 samples
sample_dat_long <- sample_data %>%
  select(Anxiety1,Anxiety2,Anxiety3,Anxiety4,Anxiety5,PID,Race) %>%
  pivot_longer(cols = c("Anxiety1","Anxiety2","Anxiety3","Anxiety4","Anxiety5"), 
               values_to = "Anxiety") %>% mutate(wave = case_when( 
                 name == "Anxiety1" ~ 1, 
                 name == "Anxiety2" ~ 2, 
                 name == "Anxiety3" ~ 3, 
                 name == "Anxiety4" ~ 4, 
                 name == "Anxiety5" ~ 5))
sample_dat_long$wave_c <- (sample_dat_long$wave) - 1
#Individual growth plots
ggplot(data = sample_dat_long, aes(x = wave_c, y = Anxiety)) + 
  geom_point(col='gray') + geom_smooth(method = "lm",col='orange') + 
  facet_wrap(vars(PID))+theme_light() 
```
### 3.b ii Individual OLS regressions conducted and visualized with the mean trajectory line.
```{r}
#Individual parametric trajectories with mean OLS trajectory
ggplot(data = sample_dat_long, aes(x = wave_c, y = Anxiety)) + 
  geom_smooth(aes(group = as.factor(PID)), method = "lm", color="gray",cex=0.8,se=F) + 
  geom_smooth(method = "lm",color ="orange",se=F,cex=0.9,lty=6)+ 
  labs(x="Wave",y="Anxiety",title="Individual parametric trajectories with mean OLS trajectory")+ 
  theme_light()
```


### 3.c i sample means of the estimated intercepts and slopes
```{r}
sample_dat_long <- sample_dat %>%
  select(Anxiety1,Anxiety2,Anxiety3,Anxiety4,Anxiety5,PID) %>% 
  pivot_longer(cols = c("Anxiety1","Anxiety2","Anxiety3","Anxiety4","Anxiety5"), 
               values_to = "Anxiety") %>% mutate(wave = case_when( 
                 name == "Anxiety1" ~ 1,
                 name == "Anxiety2" ~ 2,
                 name == "Anxiety3" ~ 3,
                 name == "Anxiety4" ~ 4,
                 name == "Anxiety5" ~ 5))
sample_dat_long$wave_c <- sample_dat_long$wave - 1

# Group by PID and create a new missing wave column
sample_dat_long_2 <- sample_dat_long %>%
  group_by(PID) %>%
  dplyr::mutate(missing_waves = sum(is.na(Anxiety)))

# Group by PID and filter for missing_wave less than 3
sample_dat_long3 <- sample_dat_long_2 %>%
  group_by(PID) %>%
  filter(sum(missing_waves) < 3)

# Building a linear model
model1 <- sample_dat_long3  %>% dplyr::group_by(PID) %>% 
  do(model = lm(Anxiety ~ wave_c, data =.))
model1[[2]][[1]]

intercept <- slope <- NULL

# Calling slope and intercept
for(i in 1:nrow(model1)){
  intercept[i] <- model1[[2]][[i]][["coefficients"]][1]
  slope[i] <- model1[[2]][[i]][["coefficients"]][2]
}
```
### 3.c ii Sample Variance
```{r}
summary(intercept)
var(intercept)
summary(slope)
var(slope)
```
### 3.c iii correlation between the estimated intercepts and slopes
```{r}
# Check the covariance of slope and intercept
cor(intercept,slope)
```

## 3.D Model building

### 3.D.i Conduct the unconditional mean model

#### 3.D.i 1 Interpret the fixed and random effects
```{r}
dat_long <- sample_dat %>%
  dplyr::select(Anxiety1,Anxiety2,Anxiety3,Anxiety4,Anxiety5,
          age1,age2,age3,age4,age5,PID)  %>% 
  pivot_longer(-PID) %>% 
  separate(name, into = c("name", "wave"), sep = "(?<=[A-Za-z])(?=[0-9])") %>% 
  pivot_wider(names_from = "name", values_from = "value")

dat_mari_race <- sample_dat %>% 
  select(Smoking, Race,PID)

dat_long <- left_join(dat_long, dat_mari_race, by = "PID")

dat_long <- remove_labels(dat_long)

dat_long$wave_c <- as.integer(dat_long$wave)-1
table(dat_long$Race)

model.a <- lmer(Anxiety ~ 1 + (1 |PID), data = dat_long, REML = FALSE)
summary(model.a)
```

#### 3.D.i 2Conduct the ICC and interpret

```{r}
performance::icc(model.a)

icc_n <- as.data.frame(VarCorr(model.a),comp="Variance")$vcov[1]
icc_d <- as.data.frame(VarCorr(model.a),comp="Variance")$vcov[1] + 
         as.data.frame(VarCorr(model.a),comp="Variance")$vcov[2]
icc_n / icc_d
```

### 3.D.ii Conduct the unconditional growth model

#### 3.D.ii 1 Interpret the fixed and random effects

```{r message=FALSE, warning=FALSE}
model.b <- lmer(Anxiety ~ wave_c + (wave_c|PID), data = dat_long, REML = FALSE)
summary(model.b)
```

#### 3.D.ii 2 Graph the unconditional growth model
```{r}
data_tmp <- data.frame(Anxiety.pred = predict(model.b),
                        wave = model.b@frame[["wave_c"]])

ggplot(data = data_tmp, mapping = aes(x = wave, y = Anxiety.pred)) +
  geom_point(col='orange',cex=0.9) +
  stat_smooth(method="lm", formula = y ~ x,cex=0.8) +
  theme_light()
```
### iii.Conduct a growth model with the main IV only

#### 1.Interpret the fixed and random effects

```{r}
table(dat_long$Race)
```

```{r}
model.c <- lmer(Anxiety ~ wave_c*Race + (wave_c|PID), data = dat_long)
summary(model.c)
```

```{r}
df.plot.c <- data.frame(Anxiety.pred = predict(model.c),
                        wave = model.c@frame[["wave_c"]],
                        Race = model.c@frame[["Race"]])

ggplot(data = df.plot.c, mapping = aes(x = wave, y = Anxiety.pred, group = Race, color = Race)) +
  geom_point(cex=0.9) +
  stat_smooth(method="lm", formula = y ~ x,cex=0.8)+theme_light()
```

### iv.Conduct a growth model with the main IV and at least one additional time-varying covariate

#### 1.Interpret the fixed and random effects

```{r message=FALSE, warning=FALSE}
model.d <- lmer(Anxiety ~ wave_c*Race + factor(age) + (wave_c|PID), data = dat_long, REML = FALSE)
summary(model.d)
```

#### 2.Graph the growth model

```{r}
plot_model(model.d, type = "pred", terms = c("wave_c", "Race", "age"))
```

### v.Using the fit statistics learned in class (i.e. Likelihood, Devianceand AIC/BIC) assess the model fit between the 4 models conducted. Which is the best model and why?

```{r message=FALSE, warning=FALSE}
df <- data.frame(fit.stats = c("-2LL", "Deviance", "AIC", "BIC"),
                 model.a = c(-2*logLik(model.a), deviance(model.a), AIC(model.a), BIC(model.a)),
                 model.b = c(-2*logLik(model.b), deviance(model.b), AIC(model.b), BIC(model.b)),
                 model.c = c(-2*logLik(model.c), deviance(model.c), AIC(model.c), BIC(model.c)),
                 model.d = c(-2*logLik(model.d), deviance(model.d), AIC(model.d), BIC(model.d)))
pander(df,caption='Model Comparison')
```


